---
title: "CanadaLogin Signal Check - Exploration Workbook"
subtitle: "Internal use only. Not a polished report."
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: true
  message: true
---

Use this workbook to keep track of queries that produce interesting metrics. If you
find something interesting, keep it in this workbook for next time. Don't remove queries
unless they no longer work or they're misleading / broken.

```{r setup}
source("R/connection.R")
source("R/colours.R")
source("R/preflight.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(glue)
  library(ggplot2)
  library(scales)
  library(ggbrick)
})

con <- connect_athena()

# Relying-party registry: application_name -> service name, operator, internal flag.
relying_parties <- load_relying_parties()
internal_apps <- relying_parties$application_name[relying_parties$is_internal]
```

## Preflight Safety Check

Run a check to ensure the data sources are up to date. Prints a checklist of each 
assumption and halts the workbook if any failed.

```{r preflight}
run_preflight_safety_check(con)
```

## Resolve the reporting period (two weeks)

Most recent complete two-week (Monday-Sunday) block that actually has sign-in
data. Graphs still plot weekly data points; this widens the period filters.

```{r resolve-period}
max_data_date <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  summarise(m = max(from_date, na.rm = TRUE)) |>
  pull(m) |>
  as.Date()

# Period end is the most recent complete Sunday: the last data day if that day is
# a Sunday, otherwise the Sunday before the data's own week. The period spans the
# two Mon-Sun weeks ending on it, so its start is 13 days earlier (a Monday).
if (as.integer(format(max_data_date, "%u")) == 7L) {
  period_end <- max_data_date
} else {
  period_end <- monday_of(max_data_date) - 1L
}
period_start <- period_end - 13L

default_period_end <- monday_of(Sys.Date()) - 1L

# Resolved period and data freshness; auto-prints as a labelled vector.
c(period_start  = format(period_start),
  period_end    = format(period_end),
  max_data_date = format(max_data_date))

if (period_end < default_period_end) {
  warning("warehouse lag: resolved period ends earlier than the default two ",
          "weeks through the most recent Sunday")
}
```

## Anomaly panel

```{r anomalies}
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(period_end)) |>
  transmute(
    day = as.Date(from_date),
    unique_users = as.double(unique_users),
    successful_logins = as.double(successful_logins),
    failed_logins = as.double(failed_logins),
    ytd_unique_users = as.double(ytd_unique_users),
    rolling_365d_unique_users = as.double(rolling_365d_unique_users)
  ) |>
  collect() |>
  arrange(day)

# Check from the actual launch date, not the launch week's Monday: Apr 20-21 are
# legitimately empty (pre-launch) and would otherwise flag as missing every week.
expected_days <- seq(as.Date(launch_date), as.Date(period_end), by = "day")
# Named list auto-prints with each element labelled by its name.
list(
  missing_days = expected_days[!expected_days %in% auth_daily$day],
  zero_success_days = auth_daily$day[auth_daily$successful_logins == 0]
)
```

## New relying parties (onboarding this period)

```{r new-rps}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  filter(first_seen >= as.Date(period_start), first_seen <= as.Date(period_end)) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(first_seen)
```

## Headline: YTD unique users (cumulative reach since launch)

The cumulative ytd column is monotonic within a week, so the week's value is its
max. (Daily `unique_users` counts user-days and must not be summed.)

```{r headline}
ytd <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  # Plot each week at its last day (Sunday week-ending date), not its start.
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

print(ytd, n = 500)

ggplot(ytd, aes(week_end, ytd_unique_users)) +
  geom_line(color = colors$canada) +
  geom_point(
    data = subset(ytd, week_end == week_end[length(week_end)]),
    color = colors$highlight[1], size = 3
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_date(breaks = ytd$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "YTD unique users", title = "YTD unique users since launch")
```

## Rolling 365-day unique users (true trailing-window reach)

`rolling_365d_unique_users` is a stored distinct-user count over the trailing 365
days, so unlike YTD it does not reset at the calendar boundary -- this is the
metric the headline should migrate to once it has enough history (and before YTD's
~Jan 2027 reset). It is null before its 2026-06-13 implementation date, so the
series starts there rather than at launch.

```{r rolling-365}
# Take each week's last available day, not the max: a trailing-365 count can fall
# once users age out past 365 days, so max() would not equal the week-ending value.
# (It currently tracks YTD because the window still reaches back before launch.)
rolling365 <- auth_daily |>
  filter(!is.na(rolling_365d_unique_users)) |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    rolling_365d_unique_users = rolling_365d_unique_users[which.max(day)],
    .groups = "drop"
  ) |>
  arrange(week_end)

print(rolling365, n = 500)

ggplot(rolling365, aes(week_end, rolling_365d_unique_users)) +
  geom_line(color = colors$canada) +
  geom_point(
    data = subset(rolling365, week_end == week_end[length(week_end)]),
    color = colors$highlight[1], size = 3
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_date(breaks = rolling365$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Rolling 365-day unique users",
       title = "Rolling 365-day unique users")
```

## Sign-in success rate (weekly)

```{r success-rate}
success_weekly <- auth_daily |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    successful_logins = sum(successful_logins, na.rm = TRUE),
    failed_logins = sum(failed_logins, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(success_rate = successful_logins / (successful_logins + failed_logins)) |>
  arrange(week_end)

print(success_weekly, n = 500)

# Y-axis is left auto-scaled, not anchored at 0: in the workbook we want to see
# the week-to-week deviations worth investigating, not a flat near-100% band.
ggplot(success_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = success_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-in success rate",
       title = "Weekly sign-in success rate since launch")
```

## Adoption breadth: sign-ins by public application (reporting period)

Internal/system apps (per the registry) excluded so the view reflects genuine
public adoption; joined to the registry for operator.

```{r sign-ins-by-app}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(
    total_logins = sum(as.double(total_logins), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")
```

### Weekly trend (since launch)

```{r sign-ins-by-app-trend}
signins_weekly <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    total_logins = as.double(total_logins)
  ) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(application_name, week_end) |>
  summarise(total_logins = sum(total_logins, na.rm = TRUE), .groups = "drop") |>
  left_join(relying_parties, by = "application_name") |>
  arrange(application_name, week_end)

print(signins_weekly, n = 500)

ggplot(signins_weekly, aes(week_end, total_logins, color = service_name)) +
  geom_line() +
  geom_point(size = 1.4) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_date(breaks = unique(signins_weekly$week_end), date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-ins", color = "Service",
       title = "Weekly sign-ins by public application since launch")
```

### Stacked weekly sign-ins by public application

Same data as above but stacked to show total volume and each app's share.
Applications below 1% of total sign-ins in the most recent week are binned
into "Other" -- the chart may therefore hide low-traffic apps.

```{r sign-ins-stacked}
latest_signin_week <- max(signins_weekly$week_end)
latest_totals <- signins_weekly |>
  filter(week_end == latest_signin_week) |>
  mutate(share = total_logins / sum(total_logins, na.rm = TRUE))
major_apps <- latest_totals$service_name[latest_totals$share >= 0.01]

launch_order <- relying_parties |>
  filter(service_name %in% major_apps) |>
  arrange(launch_date) |>
  pull(service_name) |>
  unique()
stack_levels <- c(launch_order, "Other")

signins_stacked <- signins_weekly |>
  mutate(service_label = if_else(service_name %in% major_apps, service_name, "Other")) |>
  group_by(service_label, week_end) |>
  summarise(total_logins = sum(total_logins, na.rm = TRUE), .groups = "drop") |>
  mutate(service_label = factor(service_label, levels = stack_levels))

ggplot(signins_stacked, aes(week_end, total_logins, fill = service_label)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = c(colors$primary, unname(colors$missions$dark))) +
  scale_y_continuous(labels = comma, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_x_date(breaks = unique(signins_stacked$week_end), date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-ins", fill = "Service",
       title = "Weekly sign-ins by public application (stacked)")
```

## Adoption curve by application (days since each app's own launch)

Cumulative distinct users (`ytd_unique_users`) re-based so every public app starts
at day 1 on its own launch date, capped at the first 90 days. Re-basing to
days-since-launch lines the apps up head-to-head regardless of when they onboarded,
so we can compare how quickly each builds reach in its first three months.
Internal/system apps excluded. (Apps that launched recently only extend a short way
along the x-axis; the 90-day cap is forward-looking.)

```{r adoption-curve}
adoption_curve <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  # One cumulative value per app-day (ytd is monotonic within a day, so max is safe).
  group_by(application_name, day) |>
  summarise(unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  group_by(application_name) |>
  mutate(
    # Each app's launch is its first appearance in app_login_counts; t = 1 there.
    app_launch = min(day),
    days_since_launch = as.integer(day - app_launch) + 1L
  ) |>
  ungroup() |>
  filter(days_since_launch <= 90) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(service_name, days_since_launch)

print(adoption_curve, n = 500)

ggplot(adoption_curve, aes(days_since_launch, unique_users, color = service_name)) +
  geom_line() +
  geom_point(size = 1.2) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_continuous(limits = c(1, 90), breaks = c(1, seq(15, 90, 15))) +
  theme_bw() +
  labs(x = "Days since app launch", y = "Cumulative unique users",
       color = "Service",
       title = "Adoption curve by public application",
       subtitle = "Cumulative distinct users, re-based to each app's launch day (first 90 days)")
```

## Applications per credential (cross-service reach, weekly)

Sum of each app's cumulative YTD distinct users / overall CanadaLogin YTD distinct
users (cookbook: applications-per-credential). 1.0 means the typical credential
touches one application. The cookbook flags that internal apps (migration,
profile-management) inflate this, so we show it both for all integrated apps and
for public-facing apps only.

```{r apps-per-credential}
# Per-week, take each cumulative series' week-ending value (ytd is monotonic
# within a week, so the max), then form the ratio. auth_daily already holds the
# overall YTD series; here we add the per-app sum, split by internal flag.
app_ytd_weekly <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(period_end)) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  mutate(
    week_end = monday_of(day) + 6,
    is_internal = application_name %in% internal_apps
  ) |>
  group_by(application_name, is_internal, week_end) |>
  summarise(app_ytd = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  group_by(week_end) |>
  summarise(
    sum_app_all = sum(app_ytd, na.rm = TRUE),
    sum_app_public = sum(app_ytd[!is_internal], na.rm = TRUE),
    .groups = "drop"
  )

apps_per_credential_weekly <- auth_daily |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    overall_unique_users = max(ytd_unique_users, na.rm = TRUE),
    .groups = "drop"
  ) |>
  inner_join(app_ytd_weekly, by = "week_end") |>
  mutate(
    apps_per_credential = sum_app_all / overall_unique_users,
    apps_per_credential_public = sum_app_public / overall_unique_users
  ) |>
  arrange(week_end)

print(apps_per_credential_weekly, n = 500)

apps_per_credential_weekly |>
  select(week_end,
         `All apps` = apps_per_credential,
         `Public only` = apps_per_credential_public) |>
  pivot_longer(-week_end, names_to = "scope", values_to = "apps_per_credential") |>
  ggplot(aes(week_end, apps_per_credential, color = scope)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = colors$greyscale[3]) +
  geom_line() +
  geom_point(size = 1.6) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(breaks = apps_per_credential_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Applications per credential", color = NULL,
       title = "Weekly applications per credential since launch",
       subtitle = "All integrated apps vs public-facing only (internal excluded)")
```

## MFA factor mix and success (reporting period)

"sent" rows dropped: success rate is computed from outcome events only.

```{r mfa}
tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type, result) |>
  summarise(count = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  pivot_wider(names_from = result, values_from = count, values_fill = 0) |>
  mutate(
    total = success + failure,
    success_rate = if_else(total > 0, success / total, NA_real_)
  ) |>
  arrange(desc(total))
```

### Factor mix (waffle)

```{r mfa-waffle}
# Canonical factor mix (cookbook: mfa-activity-by-factor): count "sent" for OTP
# types and outcomes for non-OTP, so OTP factors are not double-counted against
# non-OTP ones. (This is the volume-mix view; the table above carries per-factor
# success rates.)
otp_types <- c("sms_otp", "email_otp", "voice_otp")
mfa_factor_mix <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result == "sent" |
      (result %in% c("success", "failure") & !mfa_type %in% otp_types)
  ) |>
  group_by(mfa_type) |>
  summarise(mfa_auth = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  filter(mfa_auth > 0) |>
  mutate(share = mfa_auth / sum(mfa_auth)) |>
  arrange(desc(mfa_auth))

print(mfa_factor_mix, n = 500)

# Friendly OTP labels so the workbook speaks the report's language (SMS OTP /
# Email OTP / Voice OTP); non-OTP factors keep their raw mfa_type. Then round
# shares to whole percentage points and nudge the largest so the squares sum to
# exactly 100: a clean 10-per-layer waffle where one square represents 1%.
otp_display <- c(sms_otp = "SMS OTP", email_otp = "Email OTP", voice_otp = "Voice OTP")
mfa_waffle <- mfa_factor_mix |>
  mutate(
    factor_label = coalesce(unname(otp_display[mfa_type]), mfa_type),
    pct = round(share * 100)
  )
mfa_waffle$pct[which.max(mfa_waffle$pct)] <-
  mfa_waffle$pct[which.max(mfa_waffle$pct)] + (100L - sum(mfa_waffle$pct))

# geom_waffle0 (gap = 0) makes the squares touch with no white space between them;
# scale_fill_brewer pulls a ColorBrewer qualitative palette for clearer contrast.
ggplot(mfa_waffle) +
  geom_waffle0(aes(x = "MFA", y = pct, fill = factor_label), bricks_per_layer = 10) +
  coord_waffle(10) +
  scale_fill_brewer(palette = "Set1") +
  theme_void() +
  labs(
    fill = "Factor", title = "MFA factor mix this period",
    subtitle = paste0(
      "Each square represents 1% of MFA challenges ",
      "(sent for OTP, outcomes for non-OTP)"
    )
  )
```

### Weekly MFA success rate (since launch)

```{r mfa-trend}
mfa_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  transmute(day = as.Date(from_date), result, count = as.double(count)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    outcomes = sum(count, na.rm = TRUE),
    success_rate = sum(count[result == "success"], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(week_end)

print(mfa_weekly, n = 500)

ggplot(mfa_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = mfa_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "MFA success rate",
       title = "Weekly MFA success rate since launch")

# Volume of MFA outcome events per week ("sent" excluded), to read alongside the
# rate: a rate move on rising volume is a different story than on flat volume.
ggplot(mfa_weekly, aes(week_end, outcomes)) +
  geom_col(fill = colors$canada) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_date(breaks = mfa_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "MFA outcome events",
       title = "Weekly MFA outcome volume since launch")
```

## MFA success rate by factor

Per-factor success rate so we can spot whether one second factor is notably
harder to clear than another. "sent" rows excluded -- rate is from outcome
events only. Read alongside `mfa_attempts`: low-volume factors post noisy rates.

```{r mfa-success-by-factor}
mfa_by_factor <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type) |>
  summarise(
    mfa_attempts = sum(as.double(count), na.rm = TRUE),
    mfa_success_rate =
      sum(if_else(result == "success", as.double(count), 0), na.rm = TRUE) /
        sum(as.double(count), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mfa_attempts)) |>
  collect()

print(mfa_by_factor, n = 500)
```

### Weekly MFA success rate by factor (since launch)

```{r mfa-success-by-factor-trend}
mfa_factor_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  transmute(
    day = as.Date(from_date),
    mfa_type,
    result,
    count = as.double(count)
  ) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(mfa_type, week_end) |>
  summarise(
    mfa_attempts = sum(count, na.rm = TRUE),
    mfa_success_rate = sum(count[result == "success"], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(mfa_type, week_end)

print(mfa_factor_weekly, n = 500)

latest_week <- max(mfa_factor_weekly$week_end)
latest_share <- mfa_factor_weekly |>
  filter(week_end == latest_week) |>
  mutate(share = mfa_attempts / sum(mfa_attempts, na.rm = TRUE))
visible_factors <- latest_share$mfa_type[latest_share$share >= 0.005]

high_volume_factors <- mfa_factor_weekly |>
  filter(mfa_type %in% visible_factors)

ggplot(high_volume_factors, aes(week_end, mfa_success_rate, color = mfa_type)) +
  geom_line() +
  geom_point(size = 1.4) +
  scale_color_manual(values = c(colors$primary, unname(colors$missions$dark))) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = unique(mfa_factor_weekly$week_end), date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Success rate", color = "Factor",
       title = "Weekly MFA success rate by factor since launch")
```

## Passkeys (FIDO2): rollout and sign-in factor mix

Passkeys became generally available to the public on 2026-07-06 (cookbook:
sign-in-mfa-activity-by-factor). FIDO2 was already used internally before that
date, and the data cannot separate internal from public passkey users, so the
signal to watch is the change in FIDO2 volume and in its share of the public
sign-in factors once it went public. Restrict to the factors the public can
sign in with (sms_otp, voice_otp, fido2); email_otp validates emails at sign-up
and kq/totp/qr/ibm_verify_push are internal. De-duplicate per the cookbook:
count "sent" for the OTP factors and outcomes for fido2 (which has no "sent"
event), so OTP is not double-counted against passkeys.

### Weekly sign-in factor mix (since launch)

```{r signin-factor-mix}
otp_types <- c("sms_otp", "email_otp", "voice_otp")
sign_in_types <- c("sms_otp", "voice_otp", "fido2")

signin_factor_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    mfa_type %in% sign_in_types,
    result == "sent" |
      (result %in% c("success", "failure") & !mfa_type %in% otp_types)
  ) |>
  transmute(day = as.Date(from_date), mfa_type, count = as.double(count)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end, mfa_type) |>
  summarise(n = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(week_end) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  arrange(mfa_type, week_end)

print(signin_factor_weekly, n = 500)

# Weekly FIDO2 volume, so the post-2026-07-06 jump is visible against the flat
# internal-only baseline before it.
signin_factor_weekly |>
  filter(mfa_type == "fido2") |>
  ggplot(aes(week_end, n)) +
  geom_col(fill = colors$canada) +
  geom_vline(xintercept = as.Date("2026-07-06"), linetype = "dashed",
             color = colors$highlight[1]) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = unique(signin_factor_weekly$week_end),
               date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Passkey (FIDO2) sign-in events",
       title = "Weekly passkey usage since launch",
       subtitle = "Dashed line marks the 2026-07-06 public rollout")

# FIDO2 share of the public sign-in factors, week by week.
signin_factor_weekly |>
  filter(mfa_type == "fido2") |>
  ggplot(aes(week_end, share)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent, limits = c(0, NA)) +
  scale_x_date(breaks = unique(signin_factor_weekly$week_end),
               date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "FIDO2 share of public sign-in factors",
       title = "Passkey share of public sign-in second factors")
```

### Daily passkey usage around the rollout

Daily granularity (not the usual weekly points) because the rollout lands
mid-period: a weekly series collapses the step-change into a single point. FIDO2
outcome events per day; the dashed line marks the 2026-07-06 rollout.

```{r passkey-daily}
passkey_daily <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start) - 7L,
    from_date <= as.Date(period_end),
    mfa_type == "fido2",
    result %in% c("success", "failure")
  ) |>
  transmute(day = as.Date(from_date), count = as.double(count)) |>
  collect() |>
  group_by(day) |>
  summarise(passkey_signins = sum(count, na.rm = TRUE), .groups = "drop") |>
  arrange(day)

print(passkey_daily, n = 500)

ggplot(passkey_daily, aes(day, passkey_signins)) +
  geom_col(fill = colors$canada) +
  geom_vline(xintercept = as.Date("2026-07-06"), linetype = "dashed",
             color = colors$highlight[1]) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Passkey (FIDO2) sign-in events",
       title = "Daily passkey usage around the public rollout")
```

## Call centre (reporting period)

Test calls separated out; volumes are tiny, so prefer counts over rates.

```{r call-centre}
# weekly_activity_report stores its date columns as strings (catalog: "cast to
# DATE when filtering"). It is a tiny table (one row per week), so collect it and
# filter in R on parsed dates rather than wrestle Athena's date-cast translation.
tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(date_range_start >= period_start, date_range_end <= period_end) |>
  transmute(
    date_range_start,
    date_range_end,
    calls_answered = as.double(calls_answered),
    test_calls = as.double(test_calls),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  )

# weekly_topic_dump has a real DATE column, call_date. Filter on it with as.Date()
# literals: RAthena renders ISO date-like strings as Athena DATE literals, so a
# date column compares cleanly (date >= date) while the string `date` mirror
# would raise a varchar-vs-date mismatch.
tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date(period_start), call_date <= as.Date(period_end)) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count)) |>
  head(10)
```

### Weekly genuine call volume (since launch)

Volumes are tiny, so read this as direction, not precision; a topic spike is the
kind of "big change" worth chasing down.

```{r call-centre-trend}
calls_weekly <- tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    week_end = as.Date(date_range_end),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  ) |>
  filter(week_end <= period_end) |>
  transmute(week_end, genuine_calls) |>
  arrange(week_end)

print(calls_weekly, n = 500)

ggplot(calls_weekly, aes(week_end, genuine_calls)) +
  geom_col(fill = colors$canada) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = calls_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Genuine calls", title = "Weekly genuine call volume")
```

```{r teardown}
DBI::dbDisconnect(con)
```
