---
title: "CanadaLogin Fortnightly Review - Inspection Workbook"
subtitle: "Internal use only. Not a polished report."
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: true
  message: true
---

Use this workbook to keep track of queries that produce interesting metrics. If you
find something interesting, keep it in this workbook for next time. Don't remove queries
unless they no longer work or they're misleading / broken.

```{r setup}
source("R/connection.R")
source("R/colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

con <- connect_athena()

# Relying-party registry: application_name -> service name, operator, internal flag.
relying_parties <- load_relying_parties()
internal_apps <- relying_parties$application_name[relying_parties$is_internal]
```

## Resolve the reporting period (two weeks)

Most recent complete two-week (Monday-Sunday) block that actually has sign-in
data. Graphs still plot weekly data points; this widens the period filters.

```{r resolve-period}
max_data_date <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  summarise(m = max(from_date, na.rm = TRUE)) |>
  pull(m) |>
  as.Date()

# Period end is the most recent complete Sunday: the last data day if that day is
# a Sunday, otherwise the Sunday before the data's own week. The period spans the
# two Mon-Sun weeks ending on it, so its start is 13 days earlier (a Monday).
if (as.integer(format(max_data_date, "%u")) == 7L) {
  period_end <- max_data_date
} else {
  period_end <- monday_of(max_data_date) - 1L
}
period_start <- period_end - 13L

default_period_end <- monday_of(Sys.Date()) - 1L

# Resolved period and data freshness; auto-prints as a labelled vector.
c(period_start  = format(period_start),
  period_end    = format(period_end),
  max_data_date = format(max_data_date))

if (period_end < default_period_end) {
  warning("warehouse lag: resolved period ends earlier than the default two ",
          "weeks through the most recent Sunday")
}
```

## Anomaly panel

```{r anomalies}
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(period_end)) |>
  transmute(
    day = as.Date(from_date),
    unique_users = as.double(unique_users),
    successful_logins = as.double(successful_logins),
    failed_logins = as.double(failed_logins),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  arrange(day)

# Check from the actual launch date, not the launch week's Monday: Apr 20-21 are
# legitimately empty (pre-launch) and would otherwise flag as missing every week.
expected_days <- seq(as.Date(launch_date), as.Date(period_end), by = "day")
# Named list auto-prints with each element labelled by its name.
list(
  missing_days = expected_days[!expected_days %in% auth_daily$day],
  zero_success_days = auth_daily$day[auth_daily$successful_logins == 0]
)
```

## New relying parties (onboarding this period)

```{r new-rps}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  filter(first_seen >= as.Date(period_start), first_seen <= as.Date(period_end)) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(first_seen)
```

## Headline: YTD unique users (cumulative reach since launch)

The cumulative ytd column is monotonic within a week, so the week's value is its
max. (Daily `unique_users` counts user-days and must not be summed.)

```{r headline}
ytd <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  # Plot each week at its last day (Sunday week-ending date), not its start.
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

print(ytd)

ggplot(ytd, aes(week_end, ytd_unique_users)) +
  geom_line(color = colors$canada) +
  geom_point(
    data = subset(ytd, week_end == week_end[length(week_end)]),
    color = colors$highlight[1], size = 3
  ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = ytd$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "YTD unique users", title = "YTD unique users since launch")
```

## Sign-in success rate (weekly)

```{r success-rate}
success_weekly <- auth_daily |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    successful_logins = sum(successful_logins, na.rm = TRUE),
    failed_logins = sum(failed_logins, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(success_rate = successful_logins / (successful_logins + failed_logins)) |>
  arrange(week_end)

print(success_weekly)

# Y-axis is left auto-scaled, not anchored at 0: in the workbook we want to see
# the week-to-week deviations worth investigating, not a flat near-100% band.
ggplot(success_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = success_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-in success rate",
       title = "Weekly sign-in success rate since launch")
```

## Adoption breadth: sign-ins by public application (reporting period)

Internal/system apps (per the registry) excluded so the view reflects genuine
public adoption; joined to the registry for operator.

```{r sign-ins-by-app}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(total_logins = sum(as.double(total_logins), na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")
```

### Weekly trend (since launch)

```{r sign-ins-by-app-trend}
signins_weekly <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  transmute(application_name, day = as.Date(from_date), total_logins = as.double(total_logins)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(application_name, week_end) |>
  summarise(total_logins = sum(total_logins, na.rm = TRUE), .groups = "drop") |>
  left_join(relying_parties, by = "application_name") |>
  arrange(application_name, week_end)

print(signins_weekly)

ggplot(signins_weekly, aes(week_end, total_logins, color = service_name)) +
  geom_line() +
  geom_point(size = 1.4) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_date(breaks = unique(signins_weekly$week_end), date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-ins", color = "Service",
       title = "Weekly sign-ins by public application since launch")
```

## Applications per credential (cross-service reach, latest snapshot)

```{r apps-per-credential}
sum_app_unique_users <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  slice_max(from_date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  summarise(s = sum(as.double(ytd_unique_users), na.rm = TRUE)) |>
  pull(s)

overall_unique_users <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  slice_max(from_date, n = 1, with_ties = FALSE) |>
  pull(ytd_unique_users)

sum_app_unique_users / as.double(overall_unique_users)
```

## MFA factor mix and success (reporting period)

"sent" rows dropped: success rate is computed from outcome events only.

```{r mfa}
tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type, result) |>
  summarise(count = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  pivot_wider(names_from = result, values_from = count, values_fill = 0) |>
  mutate(
    total = success + failure,
    success_rate = if_else(total > 0, success / total, NA_real_)
  ) |>
  arrange(desc(total))
```

### Weekly MFA success rate (since launch)

```{r mfa-trend}
mfa_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  transmute(day = as.Date(from_date), result, count = as.double(count)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    outcomes = sum(count, na.rm = TRUE),
    success_rate = sum(count[result == "success"], na.rm = TRUE) / sum(count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(week_end)

print(mfa_weekly)

ggplot(mfa_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = mfa_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "MFA success rate",
       title = "Weekly MFA success rate since launch")
```

## Call centre (reporting period)

Test calls separated out; volumes are tiny, so prefer counts over rates.

```{r call-centre}
# weekly_activity_report stores its date columns as strings (catalog: "cast to
# DATE when filtering"). It is a tiny table (one row per week), so collect it and
# filter in R on parsed dates rather than wrestle Athena's date-cast translation.
tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(date_range_start >= period_start, date_range_end <= period_end) |>
  transmute(
    date_range_start,
    date_range_end,
    calls_answered = as.double(calls_answered),
    test_calls = as.double(test_calls),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  )

# weekly_topic_dump has a real DATE column, call_date. Filter on it with as.Date()
# literals: RAthena renders ISO date-like strings as Athena DATE literals, so a
# date column compares cleanly (date >= date) while the string `date` mirror
# would raise a varchar-vs-date mismatch.
tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date(period_start), call_date <= as.Date(period_end)) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count)) |>
  head(10)
```

### Weekly genuine call volume (since launch)

Volumes are tiny, so read this as direction, not precision; a topic spike is the
kind of "big change" worth chasing down.

```{r call-centre-trend}
calls_weekly <- tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    week_end = as.Date(date_range_end),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  ) |>
  filter(week_end <= period_end) |>
  transmute(week_end, genuine_calls) |>
  arrange(week_end)

print(calls_weekly)

ggplot(calls_weekly, aes(week_end, genuine_calls)) +
  geom_col(fill = colors$canada) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = calls_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Genuine calls", title = "Weekly genuine call volume")
```

```{r teardown}
DBI::dbDisconnect(con)
```
