---
title: "CanadaLogin Fortnightly Review - Inspection Workbook"
subtitle: "Internal use only. Not a polished report."
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: true
  message: true
---

Use this workbook to keep track of queries that produce interesting metrics. If you
find something interesting, keep it in this workbook for next time. Don't remove queries
unless they no longer work or they're misleading / broken.

```{r setup}
source("R/connection.R")
source("R/colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggbrick)
})

con <- connect_athena()

# Relying-party registry: application_name -> service name, operator, internal flag.
relying_parties <- load_relying_parties()
internal_apps <- relying_parties$application_name[relying_parties$is_internal]
```

## Resolve the reporting period (two weeks)

Most recent complete two-week (Monday-Sunday) block that actually has sign-in
data. Graphs still plot weekly data points; this widens the period filters.

```{r resolve-period}
max_data_date <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  summarise(m = max(from_date, na.rm = TRUE)) |>
  pull(m) |>
  as.Date()

# Period end is the most recent complete Sunday: the last data day if that day is
# a Sunday, otherwise the Sunday before the data's own week. The period spans the
# two Mon-Sun weeks ending on it, so its start is 13 days earlier (a Monday).
if (as.integer(format(max_data_date, "%u")) == 7L) {
  period_end <- max_data_date
} else {
  period_end <- monday_of(max_data_date) - 1L
}
period_start <- period_end - 13L

default_period_end <- monday_of(Sys.Date()) - 1L

# Resolved period and data freshness; auto-prints as a labelled vector.
c(period_start  = format(period_start),
  period_end    = format(period_end),
  max_data_date = format(max_data_date))

if (period_end < default_period_end) {
  warning("warehouse lag: resolved period ends earlier than the default two ",
          "weeks through the most recent Sunday")
}
```

## Anomaly panel

```{r anomalies}
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(period_end)) |>
  transmute(
    day = as.Date(from_date),
    unique_users = as.double(unique_users),
    successful_logins = as.double(successful_logins),
    failed_logins = as.double(failed_logins),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  arrange(day)

# Check from the actual launch date, not the launch week's Monday: Apr 20-21 are
# legitimately empty (pre-launch) and would otherwise flag as missing every week.
expected_days <- seq(as.Date(launch_date), as.Date(period_end), by = "day")
# Named list auto-prints with each element labelled by its name.
list(
  missing_days = expected_days[!expected_days %in% auth_daily$day],
  zero_success_days = auth_daily$day[auth_daily$successful_logins == 0]
)
```

## New relying parties (onboarding this period)

```{r new-rps}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  filter(first_seen >= as.Date(period_start), first_seen <= as.Date(period_end)) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(first_seen)
```

## Headline: YTD unique users (cumulative reach since launch)

The cumulative ytd column is monotonic within a week, so the week's value is its
max. (Daily `unique_users` counts user-days and must not be summed.)

```{r headline}
ytd <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  # Plot each week at its last day (Sunday week-ending date), not its start.
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

print(ytd)

ggplot(ytd, aes(week_end, ytd_unique_users)) +
  geom_line(color = colors$canada) +
  geom_point(
    data = subset(ytd, week_end == week_end[length(week_end)]),
    color = colors$highlight[1], size = 3
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_date(breaks = ytd$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "YTD unique users", title = "YTD unique users since launch")
```

## Sign-in success rate (weekly)

```{r success-rate}
success_weekly <- auth_daily |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    successful_logins = sum(successful_logins, na.rm = TRUE),
    failed_logins = sum(failed_logins, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(success_rate = successful_logins / (successful_logins + failed_logins)) |>
  arrange(week_end)

print(success_weekly)

# Y-axis is left auto-scaled, not anchored at 0: in the workbook we want to see
# the week-to-week deviations worth investigating, not a flat near-100% band.
ggplot(success_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = success_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-in success rate",
       title = "Weekly sign-in success rate since launch")
```

## Adoption breadth: sign-ins by public application (reporting period)

Internal/system apps (per the registry) excluded so the view reflects genuine
public adoption; joined to the registry for operator.

```{r sign-ins-by-app}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(
    total_logins = sum(as.double(total_logins), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")
```

### Weekly trend (since launch)

```{r sign-ins-by-app-trend}
signins_weekly <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    !application_name %in% internal_apps
  ) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    total_logins = as.double(total_logins)
  ) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(application_name, week_end) |>
  summarise(total_logins = sum(total_logins, na.rm = TRUE), .groups = "drop") |>
  left_join(relying_parties, by = "application_name") |>
  arrange(application_name, week_end)

print(signins_weekly)

ggplot(signins_weekly, aes(week_end, total_logins, color = service_name)) +
  geom_line() +
  geom_point(size = 1.4) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_date(breaks = unique(signins_weekly$week_end), date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Sign-ins", color = "Service",
       title = "Weekly sign-ins by public application since launch")
```

## Applications per credential (cross-service reach, weekly)

Sum of each app's cumulative YTD distinct users / overall CanadaLogin YTD distinct
users (cookbook: applications-per-credential). 1.0 means the typical credential
touches one application. The cookbook flags that internal apps (migration,
profile-management) inflate this, so we show it both for all integrated apps and
for public-facing apps only.

```{r apps-per-credential}
# Per-week, take each cumulative series' week-ending value (ytd is monotonic
# within a week, so the max), then form the ratio. auth_daily already holds the
# overall YTD series; here we add the per-app sum, split by internal flag.
app_ytd_weekly <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(period_end)) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  mutate(
    week_end = monday_of(day) + 6,
    is_internal = application_name %in% internal_apps
  ) |>
  group_by(application_name, is_internal, week_end) |>
  summarise(app_ytd = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  group_by(week_end) |>
  summarise(
    sum_app_all = sum(app_ytd, na.rm = TRUE),
    sum_app_public = sum(app_ytd[!is_internal], na.rm = TRUE),
    .groups = "drop"
  )

apps_per_credential_weekly <- auth_daily |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    overall_unique_users = max(ytd_unique_users, na.rm = TRUE),
    .groups = "drop"
  ) |>
  inner_join(app_ytd_weekly, by = "week_end") |>
  mutate(
    apps_per_credential = sum_app_all / overall_unique_users,
    apps_per_credential_public = sum_app_public / overall_unique_users
  ) |>
  arrange(week_end)

print(apps_per_credential_weekly)

apps_per_credential_weekly |>
  select(week_end,
         `All apps` = apps_per_credential,
         `Public only` = apps_per_credential_public) |>
  pivot_longer(-week_end, names_to = "scope", values_to = "apps_per_credential") |>
  ggplot(aes(week_end, apps_per_credential, color = scope)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = colors$greyscale[3]) +
  geom_line() +
  geom_point(size = 1.6) +
  scale_color_manual(values = colors$primary) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(breaks = apps_per_credential_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Applications per credential", color = NULL,
       title = "Weekly applications per credential since launch",
       subtitle = "All integrated apps vs public-facing only (internal excluded)")
```

## MFA factor mix and success (reporting period)

"sent" rows dropped: success rate is computed from outcome events only.

```{r mfa}
tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type, result) |>
  summarise(count = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  pivot_wider(names_from = result, values_from = count, values_fill = 0) |>
  mutate(
    total = success + failure,
    success_rate = if_else(total > 0, success / total, NA_real_)
  ) |>
  arrange(desc(total))
```

### Factor mix (waffle)

```{r mfa-waffle}
# Canonical factor mix (cookbook: mfa-activity-by-factor): count "sent" for OTP
# types and outcomes for non-OTP, so OTP factors are not double-counted against
# non-OTP ones. (This is the volume-mix view; the table above carries per-factor
# success rates.)
otp_types <- c("sms_otp", "email_otp", "voice_otp")
mfa_factor_mix <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(period_start),
    from_date <= as.Date(period_end),
    result == "sent" |
      (result %in% c("success", "failure") & !mfa_type %in% otp_types)
  ) |>
  group_by(mfa_type) |>
  summarise(mfa_auth = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  filter(mfa_auth > 0) |>
  mutate(share = mfa_auth / sum(mfa_auth)) |>
  arrange(desc(mfa_auth))

print(mfa_factor_mix)

# Round shares to whole percentage points, then nudge the largest so the squares
# sum to exactly 100: a clean 10-per-layer waffle where one square = 1% of
# challenges.
mfa_waffle <- mfa_factor_mix |>
  mutate(pct = round(share * 100))
mfa_waffle$pct[which.max(mfa_waffle$pct)] <-
  mfa_waffle$pct[which.max(mfa_waffle$pct)] + (100L - sum(mfa_waffle$pct))

# geom_waffle0 (gap = 0) makes the squares touch with no white space between them;
# scale_fill_brewer pulls a ColorBrewer qualitative palette for clearer contrast.
ggplot(mfa_waffle) +
  geom_waffle0(aes(x = "MFA", y = pct, fill = mfa_type), bricks_per_layer = 10) +
  coord_waffle(10) +
  scale_fill_brewer(palette = "Set1") +
  theme_void() +
  labs(
    fill = "Factor", title = "MFA factor mix this period",
    subtitle = "Each square = 1% of MFA challenges (sent for OTP, outcomes for non-OTP)"
  )
```

### Weekly MFA success rate (since launch)

```{r mfa-trend}
mfa_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(launch_week_start),
    from_date <= as.Date(period_end),
    result %in% c("success", "failure")
  ) |>
  transmute(day = as.Date(from_date), result, count = as.double(count)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(
    outcomes = sum(count, na.rm = TRUE),
    success_rate = sum(count[result == "success"], na.rm = TRUE) /
      sum(count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(week_end)

print(mfa_weekly)

ggplot(mfa_weekly, aes(week_end, success_rate)) +
  geom_line(color = colors$canada) +
  geom_point(color = colors$canada, size = 1.6) +
  scale_y_continuous(labels = percent) +
  scale_x_date(breaks = mfa_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "MFA success rate",
       title = "Weekly MFA success rate since launch")

# Volume of MFA outcome events per week ("sent" excluded), to read alongside the
# rate: a rate move on rising volume is a different story than on flat volume.
ggplot(mfa_weekly, aes(week_end, outcomes)) +
  geom_col(fill = colors$canada) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_date(breaks = mfa_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "MFA outcome events",
       title = "Weekly MFA outcome volume since launch")
```

## Call centre (reporting period)

Test calls separated out; volumes are tiny, so prefer counts over rates.

```{r call-centre}
# weekly_activity_report stores its date columns as strings (catalog: "cast to
# DATE when filtering"). It is a tiny table (one row per week), so collect it and
# filter in R on parsed dates rather than wrestle Athena's date-cast translation.
tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(date_range_start >= period_start, date_range_end <= period_end) |>
  transmute(
    date_range_start,
    date_range_end,
    calls_answered = as.double(calls_answered),
    test_calls = as.double(test_calls),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  )

# weekly_topic_dump has a real DATE column, call_date. Filter on it with as.Date()
# literals: RAthena renders ISO date-like strings as Athena DATE literals, so a
# date column compares cleanly (date >= date) while the string `date` mirror
# would raise a varchar-vs-date mismatch.
tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date(period_start), call_date <= as.Date(period_end)) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count)) |>
  head(10)
```

### Weekly genuine call volume (since launch)

Volumes are tiny, so read this as direction, not precision; a topic spike is the
kind of "big change" worth chasing down.

```{r call-centre-trend}
calls_weekly <- tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    week_end = as.Date(date_range_end),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  ) |>
  filter(week_end <= period_end) |>
  transmute(week_end, genuine_calls) |>
  arrange(week_end)

print(calls_weekly)

ggplot(calls_weekly, aes(week_end, genuine_calls)) +
  geom_col(fill = colors$canada) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = calls_weekly$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Genuine calls", title = "Weekly genuine call volume")
```

```{r teardown}
DBI::dbDisconnect(con)
```
