---
title: "CanadaLogin Week in Review - Inspection Workbook"
subtitle: "Internal use only. Not a polished report."
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: true
  message: true
---

Use this workbook to keep track of queries that produce interesting metrics. If you
find something interesting, keep it in this workbook for next time. Don't remove queries
unless they no longer work or they're misleading / broken.

```{r setup}
source("R/connection.R")
source("colours.R")
library(dplyr)
library(dbplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)

con <- connect_athena()

# Relying-party registry: application_name -> service name, operator, internal flag.
relying_parties <- load_relying_parties()
internal_apps <- relying_parties$application_name[relying_parties$is_internal]
```

## Resolve the reporting week

Most recent complete Monday-Sunday week that actually has sign-in data.

```{r resolve-week}
max_data_date <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  summarise(m = max(from_date, na.rm = TRUE)) |>
  pull(m) |>
  as.Date()

# Latest complete week: its Sunday is the last data day if that day is a Sunday,
# otherwise the Sunday before the data's own week.
if (as.integer(format(max_data_date, "%u")) == 7L) {
  week_end <- max_data_date
} else {
  week_end <- monday_of(max_data_date) - 1L
}
week_start <- week_end - 6L

default_sunday <- monday_of(Sys.Date()) - 1L

cat("Resolved reporting week:", format(week_start), "to", format(week_end), "\n")
cat("Max date with data:", format(max_data_date), "\n")
if (week_end < default_sunday) {
  cat("WARNING: warehouse appears to lag - resolved week is earlier than the",
      "default 'past Monday through most recent Sunday' rule.\n")
}
```

## Anomaly panel

```{r anomalies}
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date(launch_week_start), from_date <= as.Date(week_end)) |>
  transmute(
    day = as.Date(from_date),
    unique_users = as.double(unique_users),
    successful_logins = as.double(successful_logins),
    failed_logins = as.double(failed_logins),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  arrange(day)

# Check from the actual launch date, not the launch week's Monday: Apr 20-21 are
# legitimately empty (pre-launch) and would otherwise flag as missing every week.
expected_days <- seq(as.Date(launch_date), as.Date(week_end), by = "day")
cat("Missing days in auth feed since launch:\n")
print(expected_days[!expected_days %in% auth_daily$day])
cat("\nDays with zero successful logins:\n")
print(auth_daily$day[auth_daily$successful_logins == 0])
```

## New relying parties (onboarding this week)

```{r new-rps}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  filter(first_seen >= as.Date(week_start), first_seen <= as.Date(week_end)) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(first_seen)
```

## Headline: YTD unique users (cumulative reach since launch)

The cumulative ytd column is monotonic within a week, so the week's value is its
max. (Daily `unique_users` counts user-days and must not be summed.)

```{r headline}
ytd <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  arrange(week_start)

print(ytd)

ggplot(ytd, aes(week_start, ytd_unique_users)) +
  geom_line(color = colors$canada) +
  geom_point(
    data = subset(ytd, week_start == week_start[length(week_start)]),
    color = colors$highlight[1], size = 3
  ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  scale_x_date(breaks = ytd$week_start, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "YTD unique users", title = "YTD unique users since launch")
```

## Sign-in success rate (weekly)

```{r success-rate}
auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(
    successful_logins = sum(successful_logins, na.rm = TRUE),
    failed_logins = sum(failed_logins, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(success_rate = successful_logins / (successful_logins + failed_logins)) |>
  arrange(week_start)
```

## Adoption breadth: sign-ins by public application (reporting week)

Internal/system apps (per the registry) excluded so the view reflects genuine
public adoption; joined to the registry for operator.

```{r sign-ins-by-app}
tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date(week_start),
    from_date <= as.Date(week_end),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(total_logins = sum(as.double(total_logins), na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")
```

## Applications per credential (cross-service reach, latest snapshot)

```{r apps-per-credential}
sum_app_unique_users <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  slice_max(from_date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  summarise(s = sum(as.double(ytd_unique_users), na.rm = TRUE)) |>
  pull(s)

overall_unique_users <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  slice_max(from_date, n = 1, with_ties = FALSE) |>
  pull(ytd_unique_users)

sum_app_unique_users / as.double(overall_unique_users)
```

## MFA factor mix and success (reporting week)

"sent" rows dropped: success rate is computed from outcome events only.

```{r mfa}
tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date(week_start),
    from_date <= as.Date(week_end),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type, result) |>
  summarise(count = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  pivot_wider(names_from = result, values_from = count, values_fill = 0) |>
  mutate(
    total = success + failure,
    success_rate = if_else(total > 0, success / total, NA_real_)
  ) |>
  arrange(desc(total))
```

## Call centre (reporting week)

Test calls separated out; volumes are tiny, so prefer counts over rates.

```{r call-centre}
# weekly_activity_report stores its date columns as strings (catalog: "cast to
# DATE when filtering"). It is a tiny table (one row per week), so collect it and
# filter in R on parsed dates rather than wrestle Athena's date-cast translation.
tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(date_range_start >= week_start, date_range_end <= week_end) |>
  transmute(
    date_range_start,
    date_range_end,
    calls_answered = as.double(calls_answered),
    test_calls = as.double(test_calls),
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  )

# weekly_topic_dump has a real DATE column, call_date. Filter on it with as.Date()
# literals: RAthena renders ISO date-like strings as Athena DATE literals, so a
# date column compares cleanly (date >= date) while the string `date` mirror
# would raise a varchar-vs-date mismatch.
tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date(week_start), call_date <= as.Date(week_end)) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count)) |>
  head(10)
```

```{r teardown}
DBI::dbDisconnect(con)
```
