---
title: "CanadaLogin - Signal Check"
author: "EDCP Data and Research Team"  # colin.douglas@cds-snc.ca
date: "2026-06-17"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: false
    fig-width: 8
    fig-height: 3.4
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Queries below are inlined and frozen to hardcoded June 1-14, 2026 date bounds
# (the two complete Mon-Sun weeks ending on the most recent Sunday). This report
# is a point-in-time record and must reproduce identically forever, so the bounds
# and the internal-app list are literals rather than computed.
source("../R/connection.R")
source("../R/branding.R")
source("../R/colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

con <- connect_athena()

# Daily CanadaLogin-wide totals, launch through the period end. Time-series
# graphs plot weekly points even though the reporting period is two weeks, so
# this spans from the launch week, not just the period.
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date("2026-04-20"), from_date <= as.Date("2026-06-14")) |>
  transmute(
    day = as.Date(from_date),
    successful_logins = as.double(successful_logins),
    failed_logins = as.double(failed_logins),
    ytd_unique_users = as.double(ytd_unique_users),
    rolling_365d_unique_users = as.double(rolling_365d_unique_users)
  ) |>
  collect() |>
  arrange(day)

# Headline series: cumulative YTD unique users by week. The ytd column is
# monotonic within a week, so the week's value is its max. Daily unique_users
# counts user-days and must not be summed. Plot each week at its last day.
ytd_series <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

# Weekly successful first-factor events, used with the MFA successes below to
# estimate completed sign-in sessions (cookbook: completed-sign-in-sessions).
succ_weekly <- auth_daily |>
  mutate(week_start = monday_of(day), week_end = week_start + 6) |>
  group_by(week_end) |>
  summarise(succ = sum(successful_logins), fail = sum(failed_logins), .groups = "drop")

# Weekly successful MFA (second-factor) events. Outcome rows only; a "sent" OTP
# is an issuance, not a success.
mfa_success_weekly <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date("2026-04-20"),
    from_date <= as.Date("2026-06-14"),
    result == "success"
  ) |>
  transmute(day = as.Date(from_date), count = as.double(count)) |>
  collect() |>
  mutate(week_end = monday_of(day) + 6) |>
  group_by(week_end) |>
  summarise(mfa_succ = sum(count), .groups = "drop")

# Estimated completed sign-in sessions per week: one first-factor success per
# session, so successful events minus second-factor successes leaves one event
# per completed session. An estimate - see the cookbook recipe for assumptions.
sessions_series <- succ_weekly |>
  left_join(mfa_success_weekly, by = "week_end") |>
  mutate(
    mfa_succ = coalesce(mfa_succ, 0),
    est_sessions = succ - mfa_succ,
    success_rate = succ / (succ + fail)
  ) |>
  arrange(week_end)

# Relying-party registry: clean service names, operating department, internal
# flag. The internal exclusion is inlined as a frozen literal list so this
# report's analytical result cannot change if the registry is later reclassified.
relying_parties <- load_relying_parties()
internal_apps <- c(
  "Flow Application",
  "GCS Migration Solution",
  "GC Sign In - Profile Management App"
)

# Adoption breadth: sign-ins to public applications over the fortnight. Internal
# apps excluded. total_logins counts events and is summable; unique_users would
# count user-days if summed, so distinct reach uses each app's ytd figure below.
app_signins <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date("2026-06-01"),
    from_date <= as.Date("2026-06-14"),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(total_logins = sum(as.double(total_logins), na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")

# Distinct year-to-date reach per public app (latest snapshot in the window).
# ytd_unique_users is a stored distinct count, so take the latest row, not a sum.
app_ytd_reach <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(from_date <= as.Date("2026-06-14"), !application_name %in% internal_apps) |>
  group_by(application_name) |>
  slice_max(from_date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(application_name, ytd_unique_users = as.double(ytd_unique_users)) |>
  collect()

# First-ever appearance of each public application: detects new relying parties
# in the period and supplies the headline's onboarding reference lines.
app_first_seen <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  inner_join(relying_parties, by = "application_name") |>
  filter(!is_internal)

rp_launches <- app_first_seen |>
  filter(first_seen >= as.Date("2026-04-20"), first_seen <= as.Date("2026-06-14")) |>
  mutate(
    onboard_label = paste0(operator, "\njoined ", format(first_seen, "%b %d")),
    # Right-most launch labels sit to the left of their line so they do not clip.
    label_hjust = if_else(first_seen > as.Date("2026-05-20"), 1.05, -0.05)
  )

# Public relying parties first seen within the reporting fortnight, used to flag
# newly onboarded services in the adoption chart.
new_relying_parties <- app_first_seen |>
  filter(first_seen >= as.Date("2026-06-01"), first_seen <= as.Date("2026-06-14"))

# MFA factor mix over the fortnight (cookbook: mfa-activity-by-factor). OTP types
# log a "sent" plus an outcome event per challenge; non-OTP types log only an
# outcome. Count "sent" for OTP and outcomes for non-OTP so the two are not
# double-counted relative to each other.
otp_types <- c("sms_otp", "email_otp", "voice_otp")
mfa_mix <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date("2026-06-01"),
    from_date <= as.Date("2026-06-14"),
    result == "sent" |
      (result %in% c("success", "failure") & !mfa_type %in% otp_types)
  ) |>
  group_by(mfa_type) |>
  summarise(mfa_auth = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(share = mfa_auth / sum(mfa_auth)) |>
  arrange(desc(mfa_auth))

# MFA success rate over the fortnight: successes over outcome events ("sent"
# excluded, as it is not an outcome).
mfa_success_rate <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date("2026-06-01"),
    from_date <= as.Date("2026-06-14"),
    result %in% c("success", "failure")
  ) |>
  summarise(
    rate = sum(if_else(result == "success", as.double(count), 0), na.rm = TRUE) /
      sum(as.double(count), na.rm = TRUE)
  ) |>
  pull(rate)

# Call centre: weekly_activity_report stores date columns as strings (catalog:
# cast to DATE when filtering). It is tiny (one row per week), so collect and
# filter in R on parsed dates. Its weeks run Sunday-Saturday and the feed lags, so
# they do not align to the Mon-Sun reporting fortnight. We use the two most recent
# loaded call-centre weeks (May 31-Jun 6 and Jun 7-13); they end one day before
# the period's June 14, which is close enough at this volume.
call_activity <- tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(date_range_start >= as.Date("2026-05-31"), date_range_end <= as.Date("2026-06-13")) |>
  transmute(
    date_range_start, date_range_end,
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  ) |>
  arrange(date_range_start)

# weekly_topic_dump has a real DATE column, call_date; filter on it directly.
# Align the topic window to the two overlapping call-centre weeks (May 31-Jun 13)
# rather than the Mon-Sun period, so themes match the call counts above.
call_topics <- tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date("2026-05-31"), call_date <= as.Date("2026-06-13")) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count))

# Values referenced in the prose, so the text cannot drift from the queries.
latest_ytd <- ytd_series$ytd_unique_users[ytd_series$week_end == as.Date("2026-06-14")]
launch_ytd <- ytd_series$ytd_unique_users[1]
pre_period_ytd <- ytd_series$ytd_unique_users[ytd_series$week_end == as.Date("2026-05-31")]
net_new_fortnight <- latest_ytd - pre_period_ytd
rolling_latest <- tail(auth_daily$rolling_365d_unique_users[!is.na(auth_daily$rolling_365d_unique_users)], 1)

est_sessions_fortnight <- sum(sessions_series$est_sessions[sessions_series$week_end %in% as.Date(c("2026-06-07", "2026-06-14"))])
success_latest <- sessions_series$success_rate[sessions_series$week_end == as.Date("2026-06-14")]
success_prev <- sessions_series$success_rate[sessions_series$week_end == as.Date("2026-06-07")]

total_app_signins <- sum(app_signins$total_logins)
iris <- app_signins |> filter(service_name == "IRIS")
iris_reach <- app_ytd_reach$ytd_unique_users[app_ytd_reach$application_name == iris$application_name]
vac <- app_signins |> filter(str_detect(application_name, "VAC"))

sms_share <- mfa_mix$share[mfa_mix$mfa_type == "sms_otp"]
email_share <- mfa_mix$share[mfa_mix$mfa_type == "email_otp"]

cc_genuine <- sum(call_activity$genuine_calls)
# Label by the week-ending dates of the overlapping call-centre weeks (the weeks
# are Sun-Sat, so a single start-to-end span would bleed into May). Strip the
# zero-padding %d produces, e.g. "June 6 and June 13".
fmt_day <- function(d) paste0(format(d, "%B "), as.integer(format(d, "%d")))
cc_label <- paste(vapply(sort(call_activity$date_range_end), fmt_day, character(1)),
                  collapse = " and ")
```

**Reporting period: weeks of June 1-14, 2026.**

This first fortnightly CanadaLogin review covers how many people are reaching
CanadaLogin, how many sign-ins they are completing, which public services they
are using, and how reliably the service performed over the past two weeks.

Growth remained the headline. Cumulative unique users for the year reached
**`r comma(latest_ytd)`**, up **`r comma(net_new_fortnight)`** over the fortnight
and more than double the **`r comma(launch_ytd)`** recorded in the first week
after launch. The curve below shows steady, uninterrupted adoption since
CanadaLogin went live in April. This count is account-wide and includes internal
CanadaLogin usage, so it is not limited to public traffic.

```{r headline-graph}
ytd_plot <- ytd_series |>
  mutate(is_current = week_end == max(week_end))

headline_plot <- ggplot(ytd_plot, aes(week_end, ytd_unique_users)) +
  geom_vline(
    data = rp_launches, aes(xintercept = first_seen),
    color = colors$greyscale[3], linetype = "dashed", linewidth = 0.4
  ) +
  geom_text(
    data = rp_launches,
    aes(x = first_seen, y = 0.03 * max(ytd_plot$ytd_unique_users), label = onboard_label),
    hjust = rp_launches$label_hjust, vjust = 0, size = 2.8,
    color = colors$greyscale[2], lineheight = 0.9
  ) +
  geom_line(color = colors$canada, linewidth = 1) +
  geom_point(color = colors$canada, size = 1.6) +
  geom_point(
    data = subset(ytd_plot, is_current),
    color = colors$highlight[1], size = 3.2
  ) +
  geom_text(
    data = subset(ytd_plot, is_current),
    aes(label = comma(ytd_unique_users)),
    color = colors$highlight[1], vjust = -1, fontface = "bold", size = 3.6
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_date(breaks = ytd_plot$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = NULL, y = "Unique users (year to date)",
    title = "CanadaLogin reach continues to climb",
    subtitle = "Cumulative unique users for the year, by week since launch"
  )

add_cds_logo(headline_plot)
```

Behind that cumulative reach, weekly sign-in activity held steady. Using a newly
documented method, we estimate roughly **`r comma(round(est_sessions_fortnight, -1))`**
completed CanadaLogin sign-in sessions over the two weeks - a session being one
person getting all the way through the sign-in flow, however many steps it took.
On the services side, the fortnight brought
**`r comma(total_app_signins)`** sign-ins to public-facing services. **IRIS**,
run by PrairiesCan, carried the bulk with **`r comma(iris$total_logins)`**
sign-ins and has now reached **`r comma(iris_reach)`** distinct users this year.
**VAC Healthshare**, operated by Veterans Affairs Canada, added
**`r comma(vac$total_logins)`** sign-ins since joining on June 9 - the second
public service now driving steady volume.

```{r adoption-graph}
app_plot <- app_signins |>
  mutate(
    is_new = application_name %in% new_relying_parties$application_name,
    # Name and department on separate lines so labels stay narrow.
    label = paste0(service_name, if_else(is_new, " (new)", ""), "\n", operator)
  )

adoption_plot <- ggplot(app_plot, aes(reorder(label, total_logins), total_logins, fill = is_new)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = comma(total_logins)), hjust = -0.15, size = 3.6) +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = colors$canada, `TRUE` = colors$highlight[1])) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  guides(fill = "none") +
  theme_bw() +
  labs(
    x = NULL, y = "Sign-ins over the fortnight",
    title = "Sign-ins by application, weeks of June 1-14",
    subtitle = "Veterans Affairs' VAC Healthshare joined June 9"
  )

add_cds_logo(adoption_plot)
```

The service performed reliably. The weekly sign-in success rate climbed to
**`r percent(success_latest, accuracy = 1)`** in the week ending June 14, up from
**`r percent(success_prev, accuracy = 1)`** the week before and back to its
strongest readings since launch. Multi-factor authentication stayed near-universal
in success at **`r percent(mfa_success_rate, accuracy = 1)`**, with text-message
codes the dominant second factor (about **`r percent(sms_share, accuracy = 1)`** of
challenges) followed by email codes (**`r percent(email_share, accuracy = 1)`**).
The call centre handled **`r cc_genuine`** genuine calls across the two most recent
call-centre weeks (ending `r cc_label`; test calls excluded) - volumes remain very
low - with callers typically asking how to sign in to a service that uses
CanadaLogin.

```{r teardown}
DBI::dbDisconnect(con)
```
