---
title: "CanadaLogin Signal Check"
subtitle: "For the two-week period of June 15-28"
author: "EDCP Data and Research Team"  # colin.douglas@cds-snc.ca
date: "2026-06-29"
params:
  edition: 1
  ga_page_title: "CanadaLogin Signal Check #1"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: false
    fig-width: 8
    fig-height: 3.4
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Queries below are inlined and frozen to hardcoded June 15-28, 2026 date bounds
# (the two complete Mon-Sun weeks ending on the most recent Sunday). This report
# is a point-in-time record and must reproduce identically forever, so the bounds
# and the internal-app list are literals rather than computed.
source("../R/connection.R")
source("../R/branding.R")
source("../R/colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggbrick)
})

con <- connect_athena()

# Watermark date: the date the report was first created.
report_date <- as.Date("2026-06-29")

# Relying-party registry: clean service names, operating department, internal
# flag. The internal exclusion is inlined as a frozen literal list so this
# report's analytical result cannot change if the registry is later reclassified.
# This list reflects the registry on 2026-06-29 and includes the GC Digital Talent
# pilot (internal) and the migration simulator, so app-level views show only the
# public GC Digital Talent production app, not the pilot.
relying_parties <- load_relying_parties()
internal_apps <- c(
  "Flow Application",
  "GCS Migration Solution",
  "GC Sign In - Profile Management App",
  "GC Digital Talent - TBS - Pilot - Prod",
  "GCS SIC Migration RP Simulator - PRD"
)

# Headline series: system-wide cumulative YTD unique users by week, launch
# through the period end. The ytd column is monotonic within a week, so the
# week's value is its max; daily unique_users counts user-days and must not be
# summed. Time-series graphs plot weekly points even though the period is two
# weeks, so this spans from the launch week, not just the period. Plot each week
# at its last day (the Sunday week-ending date).
auth_daily <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date("2026-04-20"), from_date <= as.Date("2026-06-28")) |>
  transmute(
    day = as.Date(from_date),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  arrange(day)

ytd_series <- auth_daily |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

# Annotation lines for the headline: the two launches the reader should connect
# to the curve. CanadaLogin's own launch (Apr 22) anchors the start; GC Digital
# Talent's public production launch (Jun 25) is the new, much larger relying
# party that drives the final week's jump.
headline_launches <- data.frame(
  date = as.Date(c("2026-04-22", "2026-06-25")),
  label = c("CanadaLogin\nlaunched Apr 22", "GC Digital Talent\nlaunched Jun 25"),
  # Apr 22 sits at the far left, so its label reads to the right of the line;
  # Jun 25 sits near the right, so its label reads to the left.
  label_hjust = c(-0.04, 1.04)
)

# Adoption curve: cumulative distinct users (ytd_unique_users) re-based so every
# public app starts at day 1 on its own first appearance, capped at the first 90
# days. Re-basing lines the apps up head-to-head regardless of when they
# onboarded. Internal/system apps excluded (the literal list above), so the GC
# Digital Talent pilot is not double-counted against the public app.
adoption_curve <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date("2026-04-20"),
    from_date <= as.Date("2026-06-28"),
    !application_name %in% internal_apps
  ) |>
  transmute(
    application_name,
    day = as.Date(from_date),
    ytd_unique_users = as.double(ytd_unique_users)
  ) |>
  collect() |>
  # One cumulative value per app-day (ytd is monotonic within a day, so max is safe).
  group_by(application_name, day) |>
  summarise(unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  group_by(application_name) |>
  mutate(
    app_launch = min(day),
    days_since_launch = as.integer(day - app_launch) + 1L
  ) |>
  ungroup() |>
  filter(days_since_launch <= 90) |>
  left_join(relying_parties, by = "application_name") |>
  arrange(service_name, days_since_launch)

# MFA factor mix over the period, restricted to the two public MFA methods (SMS
# and Voice OTP). Email OTP is not true MFA. For OTP types each challenge logs a
# "sent" issuance, so "sent" is the challenge-volume measure (cookbook:
# mfa-activity-by-factor). The other factors are internal-only or in testing
# (FIDO2 security keys, push, TOTP, QR, knowledge questions) and passkeys are not
# yet released, so they are deliberately excluded from this public-method view.
otp_types <- c("sms_otp", "voice_otp")
otp_labels <- c(
  sms_otp = "SMS OTP",
  voice_otp = "Voice OTP"
)
mfa_otp_mix <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date("2026-06-15"),
    from_date <= as.Date("2026-06-28"),
    result == "sent",
    mfa_type %in% otp_types
  ) |>
  group_by(mfa_type) |>
  summarise(challenges = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(share = challenges / sum(challenges)) |>
  arrange(desc(challenges))

# Call centre: weekly_activity_report stores date columns as strings (catalog:
# cast to DATE when filtering); it is tiny (one row per week), so collect and
# filter in R on parsed dates. Its weeks run Sunday-Saturday and the feed lags, so
# they do not align to the Mon-Sun reporting period. We use the two most recent
# loaded call-centre weeks (Jun 7-13 and Jun 14-20).
call_activity <- tbl(con, in_schema("call_centre", "weekly_activity_report")) |>
  collect() |>
  mutate(
    date_range_start = as.Date(date_range_start),
    date_range_end = as.Date(date_range_end)
  ) |>
  filter(
    date_range_start >= as.Date("2026-06-07"),
    date_range_end <= as.Date("2026-06-20")
  ) |>
  transmute(
    date_range_start, date_range_end,
    genuine_calls = as.double(calls_answered) - as.double(test_calls)
  ) |>
  arrange(date_range_start)

# weekly_topic_dump has a real DATE column, call_date; filter on it directly.
# Align the topic window to the same two call-centre weeks so themes match counts.
call_topics <- tbl(con, in_schema("call_centre", "weekly_topic_dump")) |>
  filter(call_date >= as.Date("2026-06-07"), call_date <= as.Date("2026-06-20")) |>
  mutate(call_topic = str_squish(statobject)) |>
  count(call_topic, name = "topic_count") |>
  collect() |>
  mutate(topic_share = topic_count / sum(topic_count, na.rm = TRUE)) |>
  arrange(desc(topic_count))

# Values referenced in the prose, so the text cannot drift from the queries.
ytd_at <- function(week_end) {
  ytd_series$ytd_unique_users[ytd_series$week_end == as.Date(week_end)]
}
latest_ytd <- ytd_at("2026-06-28")
prev_week_ytd <- ytd_at("2026-06-21")
pre_period_ytd <- ytd_at("2026-06-14")
launch_ytd <- ytd_series$ytd_unique_users[1]
net_new_period <- latest_ytd - pre_period_ytd
net_new_final_week <- latest_ytd - prev_week_ytd

# Per-app cumulative reach at day 1 and at the latest available day, for the prose.
app_reach <- adoption_curve |>
  group_by(service_name) |>
  summarise(
    day1 = unique_users[days_since_launch == 1],
    last_day = max(days_since_launch),
    latest = unique_users[which.max(days_since_launch)],
    .groups = "drop"
  )
dt_day1 <- app_reach$day1[app_reach$service_name == "GC Digital Talent"]
dt_latest <- app_reach$latest[app_reach$service_name == "GC Digital Talent"]
iris_latest <- app_reach$latest[app_reach$service_name == "IRIS (PrairiesCan)"]
iris_days <- app_reach$last_day[app_reach$service_name == "IRIS (PrairiesCan)"]

sms_share <- mfa_otp_mix$share[mfa_otp_mix$mfa_type == "sms_otp"]
voice_share <- mfa_otp_mix$share[mfa_otp_mix$mfa_type == "voice_otp"]

cc_genuine <- sum(call_activity$genuine_calls)
# Getting-started topics: what CanadaLogin is and how to create an account.
cc_getting_started <- call_topics |>
  filter(call_topic %in% c(
    "What is CanadaLogin?", "How to Create a CanadaLogin Account"
  ))
cc_getting_started_share <- sum(cc_getting_started$topic_share)
```

Hi and welcome to the first edition of the **CanadaLogin Signal Check**, a high level 
non-technical report that highlights trends and insights from CanadaLogin metrics
over the previous two weeks. Our intention is to publish this every second week, on the
Monday after CanadaLogin sprint demo. We're still figuring out the format and cadence of 
this report, so both may shift over the next little bit. Please bear with us, and share
with us your comments and feedback.

The headline for this period of reporting is GC Digital Talent, who started using 
CanadaLogin in prod late last week. Cumulative unique users for the year reached
**`r comma(latest_ytd)`**, up **`r comma(net_new_period)`** over the two weeks, with
almost all of that, **`r comma(net_new_final_week)`** users, landing in the final week
as GC Digital Talent went live on June 25. That nearly doubled the
**`r comma(prev_week_ytd)`** we had reached the week before, and is well above the
**`r comma(launch_ytd)`** recorded in the first week after CanadaLogin launched in
April. It's important to note that this user count is system-wide, so it folds in 
internal usage. Internal unique users are a small fraction of the total, fewer than 100
users, so the headline is still valid: GC Digital Talent has doubled CanadaLogin's reach.

```{r headline-graph}
#| fig-alt: |
#|   Line chart of cumulative unique users for the year, by week, from late April
#|   to late June 2026. The curve climbs steadily near 3,000 then roughly doubles
#|   in the final week to about 7,000, with dashed markers for the CanadaLogin
#|   launch in April and the GC Digital Talent launch on June 25.

ytd_plot <- ytd_series |>
  mutate(is_current = week_end == max(week_end))

headline_plot <- ggplot(ytd_plot, aes(week_end, ytd_unique_users)) +
  geom_vline(
    data = headline_launches, aes(xintercept = date),
    color = colors$greyscale[3], linetype = "dashed", linewidth = 0.4
  ) +
  geom_text(
    data = headline_launches,
    aes(x = date, y = 0.97 * max(ytd_plot$ytd_unique_users), label = label),
    hjust = headline_launches$label_hjust, vjust = 1, size = 2.8,
    color = colors$greyscale[2], lineheight = 0.9
  ) +
  geom_line(color = colors$brand["blue"], linewidth = 1) +
  geom_point(color = colors$brand["blue"], size = 1.6) +
  geom_point(
    data = subset(ytd_plot, is_current),
    color = colors$brand["red"], size = 3.2
  ) +
  geom_text(
    data = subset(ytd_plot, is_current),
    aes(label = comma(ytd_unique_users)),
    color = colors$brand["red"], vjust = -1, fontface = "bold", size = 3.6
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_date(breaks = ytd_plot$week_end, date_labels = "%b %d") +
  theme_cds() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = NULL, y = "Unique users (year to date)",
    title = "GC Digital Talent doubles CanadaLogin's reach",
    subtitle = "Cumulative unique users for the year, by week since launch"
  )

add_cds_logo(add_watermark(headline_plot, date = report_date, edition = params$edition))
```

The next chart lines up each relying party from its own launch day. GC Digital
Talent is head and shoulders above the past two partners onboarded, in terms of
user-base. On its first day it reached **`r comma(dt_day1)`** distinct users and
**`r comma(dt_latest)`** within four days, already more than **IRIS (PrairiesCan)** has
reached in its first **`r iris_days`** days (**`r comma(iris_latest)`**). IRIS and VAC Healthshare
built their user bases gradually over weeks, whereas GC Digital Talent arrived all at 
once. The change reflects the Digital Talent team's communication approach; they emailed
their (relatively tech-savvy) userbase to let them know on launch day.

```{r adoption-graph}
#| fig-alt: |
#|   Line chart of cumulative distinct users against days since each public
#|   service's own launch, over the first 90 days. GC Digital Talent rises almost
#|   vertically past 2,700 users in four days, far steeper than IRIS (PrairiesCan) and VAC
#|   Healthshare, which climb gradually over weeks.

adoption_origin <- adoption_curve |>
  distinct(service_name) |>
  mutate(days_since_launch = 0L, unique_users = 0)

adoption_plot <- ggplot(
  bind_rows(adoption_origin, adoption_curve),
  aes(days_since_launch, unique_users, color = service_name)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.2) +
  scale_color_manual(values = c(
    "GC Digital Talent" = unname(colors$brand_action["orange"]),
    "IRIS (PrairiesCan)" = unname(colors$brand["blue"]),
    "VAC Healthshare" = unname(colors$brand_base["forest"])
  )) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_continuous(limits = c(0, 90), breaks = c(0, seq(15, 90, 15))) +
  theme_cds() +
  labs(
    x = "Days since the service launched", y = "Cumulative unique users",
    color = NULL,
    title = "A bigger service from day one",
    subtitle = paste0(
      "Cumulative distinct users, re-based to each service's launch day ",
      "(first 90 days)"
    )
  )

add_cds_logo(add_watermark(adoption_plot, date = report_date, edition = params$edition))
```

A new userbase this size is a good moment to look at how people prove who they are.
The two MFA methods available to the public are SMS OTP and Voice OTP. SMS dominates at
**`r percent(sms_share, accuracy = 1)`** of challenges, with Voice OTP making up the
remaining **`r percent(voice_share, accuracy = 1)`**. We are setting this baseline now
because passkeys, a faster and more phishing-resistant way to sign in, are releasing
soon; the waffle below is the "before" picture we will measure that shift against.

```{r mfa-graph}
#| fig-alt: |
#|   Waffle chart of 100 squares showing the mix of MFA methods available to the
#|   public. SMS OTP fills the large majority of the grid, with Voice OTP
#|   occupying a small slice.

# Round shares to whole percentage points, then nudge the largest so the squares
# sum to exactly 100: a clean 10-per-layer waffle where one square = 1% of these
# OTP challenges.
mfa_waffle <- mfa_otp_mix |>
  mutate(
    method = unname(otp_labels[mfa_type]),
    pct = round(share * 100)
  )
mfa_waffle$pct[which.max(mfa_waffle$pct)] <-
  mfa_waffle$pct[which.max(mfa_waffle$pct)] + (100L - sum(mfa_waffle$pct))

brand_font <- if (register_cds_fonts()) cds_font else ""

# ggbrick's geom_waffle0 needs a non-factor fill (a factor trips its discrete
# scale), so method is a plain character and `breaks` fixes the legend order.
mfa_plot <- ggplot(mfa_waffle) +
  geom_waffle0(aes(x = "MFA", y = pct, fill = method), bricks_per_layer = 10) +
  coord_waffle(10) +
  scale_fill_manual(
    breaks = unname(otp_labels),
    values = c(
      "SMS OTP" = unname(colors$brand["blue"]),
      "Voice OTP" = unname(colors$brand_action["orange"])
    )
  ) +
  theme_void(base_family = brand_font) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30"),
    legend.title = element_blank()
  ) +
  labs(
    fill = NULL,
    title = "How people verify today, before passkeys arrive",
    subtitle = "Each square represents 1% of MFA challenges over the two weeks"
  )

add_cds_logo(add_watermark(mfa_plot, date = report_date, edition = params$edition))
```

The call centre stayed quiet through the period, handling **`r cc_genuine`** genuine
calls across the two most recent call-centre weeks (ending June 13 and June 20; test
calls excluded), so the volume is too small to read much into. Callers were mostly
asking the basics, what CanadaLogin is and how to set up an account
(**`r percent(cc_getting_started_share, accuracy = 1)`** of topics between them).

```{r teardown}
DBI::dbDisconnect(con)
```

{{< include _footer.qmd >}}

{{< include _analytics.qmd >}}
