---
title: "CanadaLogin - Week in Review"
author: "EDCP Data and Research Team"  # colin.douglas@cds-snc.ca
date: "2026-06-17"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: false
    fig-width: 8
    fig-height: 3.4
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Queries below are inlined and frozen to hardcoded June 8-14, 2026 date bounds:
# this report is a point-in-time record and must reproduce identically forever.
source("../R/connection.R")
source("../colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(dbplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

con <- connect_athena()

# Headline: weekly snapshot of cumulative YTD unique users, launch through the
# reporting week. The cumulative column is monotonic within a week, so the
# week's value is its max. Daily unique_users counts user-days and is not summed.
ytd_series <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date("2026-04-20"), from_date <= as.Date("2026-06-14")) |>
  transmute(day = as.Date(from_date), ytd_unique_users = as.double(ytd_unique_users)) |>
  collect() |>
  mutate(week_start = monday_of(day)) |>
  group_by(week_start) |>
  summarise(ytd_unique_users = max(ytd_unique_users, na.rm = TRUE), .groups = "drop") |>
  # Plot each week at its LAST day (the Sunday week-ending date), not its start.
  mutate(week_end = week_start + 6) |>
  arrange(week_end)

# Relying-party registry: clean service names, operating department, and the
# internal-app flag. The internal exclusion below is also inlined as a frozen
# literal list so this report's analytical result cannot change if the registry
# is later reclassified; the registry is used for display labels and operator.
relying_parties <- load_relying_parties()
internal_apps <- c(
  "Flow Application",
  "GCS Migration Solution",
  "GC Sign In - Profile Management App"
)

# Adoption breadth: sign-ins by public application this week. Internal/system
# apps are excluded so the view reflects genuine public adoption.
app_signins <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  filter(
    from_date >= as.Date("2026-06-08"),
    from_date <= as.Date("2026-06-14"),
    !application_name %in% internal_apps
  ) |>
  group_by(application_name) |>
  summarise(
    total_logins = sum(as.double(total_logins), na.rm = TRUE),
    unique_users = sum(as.double(unique_users), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_logins)) |>
  collect() |>
  left_join(relying_parties, by = "application_name")

# First-ever appearance of each public application, used both to detect new
# relying parties this week and to annotate the headline with onboarding dates.
app_first_seen <- tbl(con, in_schema("ibm_verify", "app_login_counts")) |>
  group_by(application_name) |>
  summarise(first_seen = min(from_date, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(first_seen = as.Date(first_seen)) |>
  inner_join(relying_parties, by = "application_name") |>
  filter(!is_internal)

new_relying_parties <- app_first_seen |>
  filter(first_seen >= as.Date("2026-06-08"), first_seen <= as.Date("2026-06-14"))

# Onboarding dates within the headline window, for the reference-line labels.
rp_launches <- app_first_seen |>
  filter(first_seen >= as.Date("2026-04-20"), first_seen <= as.Date("2026-06-14")) |>
  mutate(
    onboard_label = paste0(operator, "\njoined ", format(first_seen, "%b %d")),
    # Right-most launch labels sit to the left of their line so they do not clip.
    label_hjust = if_else(first_seen > as.Date("2026-05-20"), 1.05, -0.05)
  )

# Service health: weekly sign-in success rate and MFA adoption.
success_week <- tbl(con, in_schema("ibm_verify", "auth_total_logins")) |>
  filter(from_date >= as.Date("2026-06-08"), from_date <= as.Date("2026-06-14")) |>
  summarise(
    successful_logins = sum(as.double(successful_logins), na.rm = TRUE),
    failed_logins = sum(as.double(failed_logins), na.rm = TRUE)
  ) |>
  collect() |>
  mutate(success_rate = successful_logins / (successful_logins + failed_logins))

mfa_week <- tbl(con, in_schema("ibm_verify", "mfa_activity")) |>
  filter(
    from_date >= as.Date("2026-06-08"),
    from_date <= as.Date("2026-06-14"),
    result %in% c("success", "failure")
  ) |>
  group_by(mfa_type, result) |>
  summarise(count = sum(as.double(count), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  pivot_wider(names_from = result, values_from = count, values_fill = 0) |>
  mutate(total = success + failure) |>
  arrange(desc(total))

# Values referenced in the prose, so the text cannot drift from the queries.
latest_ytd <- ytd_series$ytd_unique_users[nrow(ytd_series)]
launch_ytd <- ytd_series$ytd_unique_users[1]
prev_ytd <- ytd_series$ytd_unique_users[nrow(ytd_series) - 1]
net_new_users <- latest_ytd - prev_ytd
success_rate_pct <- round(100 * success_week$success_rate)
mfa_total <- sum(mfa_week$total)
mfa_success_pct <- round(100 * sum(mfa_week$success) / mfa_total)
top_factor_share <- round(100 * mfa_week$total[1] / mfa_total)
vac <- app_signins |> filter(str_detect(application_name, "VAC"))
iris <- app_signins |> filter(service_name == "IRIS")
```

**Reporting period: June 8 to 14, 2026.**

This CanadaLogin Week in Review covers how many people are using CanadaLogin,
which services they are signing in to, and how reliably the service performed
over the week.

The headline remains growth. Cumulative unique users for the year reached
**`r comma(latest_ytd)`**, up **`r comma(net_new_users)`** from the week before
and more than double the **`r comma(launch_ytd)`** recorded in the first week
after launch. The trend below shows steady, uninterrupted adoption since
CanadaLogin went live in April.

```{r headline-graph}
ytd_plot <- ytd_series |>
  mutate(is_current = week_end == max(week_end))

ggplot(ytd_plot, aes(week_end, ytd_unique_users)) +
  geom_vline(
    data = rp_launches, aes(xintercept = first_seen),
    color = colors$greyscale[3], linetype = "dashed", linewidth = 0.4
  ) +
  geom_text(
    data = rp_launches,
    aes(x = first_seen, y = 0.5 * max(ytd_plot$ytd_unique_users), label = onboard_label),
    hjust = rp_launches$label_hjust, vjust = 1, size = 2.8,
    color = colors$greyscale[2], lineheight = 0.9
  ) +
  geom_line(color = colors$canada, linewidth = 1) +
  geom_point(color = colors$canada, size = 1.6) +
  geom_point(
    data = subset(ytd_plot, is_current),
    color = colors$highlight[1], size = 3.2
  ) +
  geom_text(
    data = subset(ytd_plot, is_current),
    aes(label = comma(ytd_unique_users)),
    color = colors$highlight[1], vjust = -1, fontface = "bold", size = 3.6
  ) +
  scale_y_continuous(
    labels = comma, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_date(breaks = ytd_plot$week_end, date_labels = "%b %d") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = NULL, y = "Unique users (year to date)",
    title = "CanadaLogin reach continues to climb",
    subtitle = "Cumulative unique users for the year, by week since launch"
  )
```

The big news this week is a new service joining CanadaLogin. **VAC Healthshare**,
operated by Veterans Affairs Canada, onboarded on June 9 and was active
immediately, drawing **`r comma(vac$total_logins)`** sign-ins from
**`r comma(vac$unique_users)`** users in its first few days. It joins IRIS, run by
PrairiesCan, as the second public-facing service driving sign-in volume this
week. Each new relying party widens the base of Canadians who reach government
services through a single CanadaLogin credential.

```{r adoption-graph}
app_plot <- app_signins |>
  mutate(
    is_new = application_name %in% new_relying_parties$application_name,
    # Name and department on separate lines so labels stay narrow.
    label = paste0(service_name, if_else(is_new, " (new)", ""), "\n", operator)
  )

ggplot(app_plot, aes(reorder(label, total_logins), total_logins, fill = is_new)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = comma(total_logins)), hjust = -0.15, size = 3.6) +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = colors$canada, `TRUE` = colors$highlight[1])) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  guides(fill = "none") +
  theme_bw() +
  labs(
    x = NULL, y = "Sign-ins this week",
    title = "Sign-ins by application, week of June 8-14",
    subtitle = "Veterans Affairs' VAC Healthshare joined this week"
  )
```

Service performance held up well as the new traffic arrived. The overall sign-in
success rate was **`r success_rate_pct`%** for the week, an improvement on recent
weeks, and multi-factor authentication continued to protect accounts:
**`r comma(mfa_total)`** second-factor checks completed at a
**`r mfa_success_pct`%** success rate, with text-message codes the most common
method (**`r top_factor_share`%** of checks). *Call centre figures for this week
are not yet available (the latest loaded reporting week ends June 6) and will
appear in a future review.*

```{r teardown}
DBI::dbDisconnect(con)
```
