---
title: "CanadaLogin Signal Check" 
subtitle: "For the two-week period of January 1-14"
author: "EDCP Data and Research Team"  # example@cds-snc.ca
date: "2026-01-01"
params:
  edition: 0
  ga_page_title: "Signal Check #0"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: false
    fig-width: 8
    fig-height: 3.4
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("../R/branding.R")
source("../R/colours.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
})

report_date <- as.Date("2026-01-01")
```

This is a style template for the biweekly Signal Check. The graphs below
demonstrate the report's visual conventions: brand colours, CDS logo placement,
watermark, `theme_cds()` typography, and axis formatting.

Each body paragraph opens with a bolded action title: a single-clause statement
of the paragraph's takeaway that sets the reader's expectation before they read
on. It does not need to flow grammatically into the sentence that follows.

**The headline is a called-out time series.** The first graph is a time-series
line chart, the format used for the headline metric each report. The most recent
data point is called out in brand red with a bold label.

```{r headline-graph}
#| fig-alt: |
#|   Line chart of US personal savings rate from 1967 to 2015, used as a
#|   placeholder for the headline unique-users growth curve. The final data
#|   point is highlighted in red.

econ <- economics |>
  filter(date >= as.Date("2010-01-01")) |>
  mutate(is_current = date == max(date))

headline_plot <- ggplot(econ, aes(date, psavert)) +
  geom_line(color = colors$brand["blue"], linewidth = 1) +
  geom_point(
    data = subset(econ, is_current),
    color = colors$brand["red"], size = 3.2
  ) +
  geom_text(
    data = subset(econ, is_current),
    aes(label = paste0(psavert, "%")),
    color = colors$brand["red"], vjust = -1, fontface = "bold", size = 3.6
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  theme_cds() +
  labs(
    x = NULL, y = "Personal savings rate (%)",
    title = "Placeholder headline metric over time",
    subtitle = "US personal savings rate (economics dataset), 2010 onward"
  )

add_cds_logo(add_watermark(headline_plot, date = report_date, edition = params$edition))
```

**Bar charts line up categories for comparison.** The second graph is a
horizontal bar chart, the format used for comparing categories side by side. One
bar is highlighted in brand orange to show how a new or notable entry can be
called out.

```{r adoption-graph}
#| fig-alt: |
#|   Horizontal bar chart of vehicle count by manufacturer from the mpg
#|   dataset, used as a placeholder for sign-ins by application. Bars are
#|   coloured in brand blue, with one highlighted in orange as a new entry.

top_makes <- mpg |>
  count(manufacturer, name = "n_models") |>
  slice_max(n_models, n = 8) |>
  mutate(highlight = manufacturer == "audi")

adoption_plot <- ggplot(
  top_makes,
  aes(reorder(manufacturer, n_models), n_models, fill = highlight)
) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_models), hjust = -0.15, size = 3.6) +
  coord_flip() +
  scale_fill_manual(
    values = c("FALSE" = unname(colors$brand["blue"]), "TRUE" = unname(colors$brand_action["orange"]))
  ) +
  scale_y_continuous(
    limits = c(0, NA), expand = expansion(mult = c(0, 0.18))
  ) +
  guides(fill = "none") +
  theme_cds() +
  labs(
    x = NULL, y = "Vehicle count",
    title = "Placeholder adoption by category",
    subtitle = "Top manufacturers by model count (mpg dataset)"
  )

add_cds_logo(add_watermark(adoption_plot, date = report_date, edition = params$edition))
```

**Dense weekly series stress-test the date axis.** The third graph is a dense
weekly time series. It shows how `theme_cds()` handles a long date axis with many
data points. The line uses brand forest green, a second single-series colour
option beyond the headline blue.

```{r trend-graph}
#| fig-alt: |
#|   Line chart of average diamond price by week from 2008 to 2014, used as a
#|   placeholder for a weekly sign-in success rate trend. The line is coloured
#|   in brand forest green.

diamond_weekly <- diamonds |>
  mutate(
    price_bin = cut(carat, breaks = seq(0, 6, 0.5)),
    week = as.Date(cut(as.Date("2008-01-01") + row_number() %% (365 * 6), "week"))
  ) |>
  group_by(week) |>
  summarise(avg_price = mean(price), .groups = "drop") |>
  filter(!is.na(week))

trend_plot <- ggplot(diamond_weekly, aes(week, avg_price)) +
  geom_line(color = colors$brand_base["forest"], linewidth = 1) +
  geom_point(color = colors$brand_base["forest"], size = 1.2) +
  scale_y_continuous(
    labels = dollar, limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_cds() +
  labs(
    x = NULL, y = "Average price ($)",
    title = "Placeholder weekly trend",
    subtitle = "Average diamond price by week (diamonds dataset)"
  )

add_cds_logo(add_watermark(trend_plot, date = report_date, edition = params$edition))
```

Reports typically close with a short one-liner that sums up any low-volume
supplementary data the report did not graph above.

{{< include _footer.qmd >}}

{{< include _analytics.qmd >}}
